Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-15 ind_files/04_field_eqns.tex
Correct we/us to I/me for consistency

If you know and care about enough mathematics for this to really bother you, see~\citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes we study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions~\citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see~\cite{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I adopt this attitude below and hope that all questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future. Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, the stochastic process we will formulate $\{\phi^{(t)}\}_{t \geq 0}$ when viewed as a sequence of functions $\{\phi^{(t)}(y)\}_{t \geq 0}$ thus describes a (scalar) `stochastic field', and the formalism we will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function~\citep{bressloff_stochastic_2010} and collective motion~\citep{o_laighleis_minimal_2018}. In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$

@@

To proceed, as before, we assume that there exists a system-size parameter $K > 0$ to obtain a new process $\{\phi^{(t)}\}_{t \geq 0}$ such that for any set $A \subset \mathcal{T}$, $\int_A\phi^{(t)}dx$ gives the `density' of individuals that have trait values that lie within the set $A$. Note that we expect this stochastic process to evolve continuously if $K$ is large since the contribution of each individual is negligible. Specifically, we assume that there exists a $K>0$ such that we can make the substitutions:
\begin{align*}
    \phi^{(t)} = \frac{1}{K}\nu^{(t)} &= \frac{1}{K}\sum\limits_{i=1}^{N(t)}\delta_{x_i}\\
    b_K(x|\phi^{(t)}) &= \frac{1}{K}b(x|\nu^{(t)})\\
    d_K(x|\phi^{(t)}) &= \frac{1}{K}d(x|\nu^{(t)})
\end{align*}
$\{\phi^{(t)}\}_{t\geq0}$ takes values in 
\begin{equation*}
    \mathcal{M}_{K}(\mathcal{T}) = \left\{\frac{1}{K}\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}
\end{equation*}
In terms of these new variables, we obtain the master equation:
\begin{equation}
\label{M_equation}
    \frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[(\Delta^{-}_{x}-1)b_K(x|\phi)P(\phi,t) +(\Delta^{+}_{x}-1)d_K(x|\phi)P(\phi,t)\right]dx
\end{equation}
where we have introduced new step operators $\Delta_{x}^{\pm}$ that satisfy:
\begin{equation*}
    \Delta_{x}^{\pm}[F(y,\phi)] =  F\left(y,\phi \pm \frac{1}{K}\delta_x\right)
\end{equation*}

@@

This constitutes the `mesoscopic' description of our process. We can once again appeal to the link between Fokker-Planck equations and Langevins~\citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) for finite-dimensional stochastic processes is part of the standard mathematical canon, the corresponding equivalence is much less well understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear for general measure spaces - We may need $\mathcal{M}_{K}(\mathcal{T})$ to have a considerable amount of structure (Ex: Seperable Hilbert space) for things to work out~\citep{da_prato_stochastic_2014,balan_gentle_2018}, and I do not know whether our domain is nice enough. For example, it is not immediately clear what conditions on $\mathcal{A}^{\pm}(x | \phi)$ are needed for \eqref{functional_langevin} to even admit a solution. Nevertheless, we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, and under certain assumptions, equations like \eqref{nD_eqn_for_frequencies} have been proved to exist.

@@

where we have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE models' discussed in studies of Adaptive Diversification~\citep{doebeli_adaptive_2011}. Equation \eqref{deterministic_traj} is also the basic equation of `oligomorphic dynamics'~\citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology~\citep{nordbotten_dynamics_2020}. A prominent recent example is the `trait space equations' of~\citep{wickman_theoretical_2022} in their framework for eco-evolutionary community dynamics. As before, I assume that the birth and death functions take the form:

@@

where we have defined $w(x|\psi) \coloneqq b^{\textrm{(ind)}}(x|\psi) - d^{\textrm{(ind)}}(x|\psi)$, which can be thought of as the (Malthusian) `fitness' of the phenotype $x$. To track population numbers and trait frequencies, we follow the approach of~\citep{week_white_2021} and define

@@

Where we have used the definition of $N_K(t)$ and assumed that integrals and derivatives commute in the second line. Substituting \eqref{PDE_for_kimura}, we now obtain

@@

where we have used the definition of $p(x,t)$ in the second line. Using \eqref{mean_fitness_for_kimura} and rearranging the terms gives us:

@@

This is a continuous version of the replicator-mutator equation when each $x$ is viewed as a strategy. It also yields Kimura's continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that we have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb

@@

where we have used the Fubini-Tonnelli theorem to go from the first step to the second, and have made the substitution $u = y-z$ to go from the third to the fourth step. We then note that since $m$ is a kernel, it satisfies $\int\limits_{\mathbb{R}} m(u)du = 1$, and \eqref{convolution_integral_intermediate} therefore becomes $\int\limits_{\mathbb{R}} Q(y|\psi)dy = N_K(t)$. Substituting this in \eqref{cts_replicator_mutator}, we have

@@

is the statistical covariance of the quantity $f$ with the Malthusian fitness function (Importantly, just like in the Price equation, this is an \emph{analogy} - Everything here is deterministic). The second term, which we will denote by

@@

Under these assumptions, if we supply an initial condition $\psi(x,0) = N_{K}(0)\delta_{y_0}$ for some constants $N_K(0) > 0$ and $y_0 \in \mathcal{T}$ (meaning we start with a completely monomorphic population of size $N_K(0)$ in which all individuals have trait value $y_0$), then it is reasonable to assume that the population remains sufficiently clustered for some (possibly small) time $t>0$ that we can continue to approximate the distribution $\psi(x,t)$ as a Dirac Delta mass $N_{K}(t)\delta_{y(t)}$ that is moving across the trait space in a deterministic manner dictated by a function $y(t)$ (to be found). Note that we have $p(x,t) = \delta_{y(t)}$, $\overline{x}(t) = y(t)$, and $\overline{w}(t) = 0$. Thus, from equation \eqref{cts_price}, we have

@@

where we have used $\overline{x}(t) = y(t)$. We can define the shorthand $B(y) =\int\limits_{\mathcal{T}}(x-y(t))^2p(x,t)dx = \int\limits_{\mathcal{T}}(x-\overline{x}(t))^2p(x,t)dx$ to obtain:

@@

\section{Stochastic trait frequency dynamics in the infinite-dimensional case}
In chapter \ref{chap_BD}, we also derived SDEs for the trait frequency dynamics of the complete stochastic case using It\^o's formula. Doing the same for quantitative traits in our framework is tricky because it requires us to find a version of It\^o's formula that holds for SPDEs of the form \eqref{functional_langevin}. The formulation of It\^o formulas and/or an infinite-dimensional stochastic calculus for general function (measure) valued stochastic processes is an active area of research in pure mathematics~\citep{da_prato_stochastic_2014}, and without further information on the nature of the domain $\mathcal{M}_{K}(\mathcal{T})$, it is not clear (to me at least) whether an It\^o's formula exists for our case in general. However, it bears noting that the `intuitive' It\^o's formula one would expect does indeed hold true for broad classes of Hilbert space valued processes~\citep{prevot_concise_2007,da_prato_stochastic_2014,liu_stochastic_2015}. If $\mathcal{A}^{\pm}(x | \phi)$ in \eqref{functional_FPE} are Gaussian functions, then it has been proven~\citep{week_white_2021} that the SDEs `work as expected' if we take $m \to \infty$ in \eqref{nD_eqn_for_frequencies} (see section \ref{sec_fun_theorems} in chapter \ref{chap_unification}). However, carrying out a general derivation is beyond the scope of this thesis.

@@

We can also formally carry out a functional analogue of the weak noise expansion as we did in chapter \ref{chap_BD}. Assume that $\psi(x,t)$ is the deterministic trajectory obtained as the solution to \eqref{deterministic_traj}. We introduce a new process $\{\zeta^{(s)}\}_{s \geq 0}$ which measures the fluctuations of $\phi^{(t)}$ from the deterministic trajectory $\psi(x,t)$. More precisely, we introduce the new variables:

>>>>

If you know and care about enough mathematics for this to really bother you, see~\citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes I study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions~\citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see~\cite{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I adopt this attitude below and hope that all questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future. Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, the stochastic process I will formulate when viewed as a sequence of functions thus describes a (scalar) `stochastic field', and the formalism I will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function~\citep{bressloff_stochastic_2010} and collective motion~\citep{o_laighleis_minimal_2018}. In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$

@@

To proceed, as before, I assume that there exists a system-size parameter $K > 0$ to obtain a new process $\{\phi^{(t)}\}_{t \geq 0}$ such that for any set $A \subset \mathcal{T}$, $\int_A\phi^{(t)}dx$ gives the `density' of individuals that have trait values that lie within the set $A$. Note that we expect this stochastic process to evolve continuously if $K$ is large since the contribution of each individual is negligible. Specifically, I assume that there exists a $K>0$ such that we can make the substitutions:
\begin{align*}
    \phi^{(t)} = \frac{1}{K}\nu^{(t)} &= \frac{1}{K}\sum\limits_{i=1}^{N(t)}\delta_{x_i}\\
    b_K(x|\phi^{(t)}) &= \frac{1}{K}b(x|\nu^{(t)})\\
    d_K(x|\phi^{(t)}) &= \frac{1}{K}d(x|\nu^{(t)})
\end{align*}
$\{\phi^{(t)}\}_{t\geq0}$ takes values in 
\begin{equation*}
    \mathcal{M}_{K}(\mathcal{T}) = \left\{\frac{1}{K}\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}
\end{equation*}
In terms of these new variables, we obtain the master equation:
\begin{equation}
\label{M_equation}
    \frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[(\Delta^{-}_{x}-1)b_K(x|\phi)P(\phi,t) +(\Delta^{+}_{x}-1)d_K(x|\phi)P(\phi,t)\right]dx
\end{equation}
where I have introduced new step operators $\Delta_{x}^{\pm}$ that satisfy:
\begin{equation*}
    \Delta_{x}^{\pm}[F(y,\phi)] =  F\left(y,\phi \pm \frac{1}{K}\delta_x\right)
\end{equation*}

@@

This constitutes the `mesoscopic' description of our process. We can once again appeal to the link between Fokker-Planck equations and Langevins~\citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) for finite-dimensional stochastic processes is part of the standard mathematical canon, the corresponding equivalence is much less well understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear for general measure spaces - We may need $\mathcal{M}_{K}(\mathcal{T})$ to have a considerable amount of structure (Ex: Seperable Hilbert space) for things to work out~\citep{da_prato_stochastic_2014,balan_gentle_2018}, and I do not know whether our domain is nice enough. For example, it is not immediately clear what conditions on $\mathcal{A}^{\pm}(x | \phi)$ are needed for \eqref{functional_langevin} to even admit a solution. Nevertheless, as I show below, we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, illustrating consistency with known models, and under certain assumptions, solutions to equations like \eqref{nD_eqn_for_frequencies} have been rigorously proven to exist.

@@

where I have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE models' discussed in studies of Adaptive Diversification~\citep{doebeli_adaptive_2011}. Equation \eqref{deterministic_traj} is also the basic equation of `oligomorphic dynamics'~\citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology~\citep{nordbotten_dynamics_2020}. A prominent recent example is the `trait space equations' of~\citep{wickman_theoretical_2022} in their framework for eco-evolutionary community dynamics. As before, I assume that the birth and death functions take the form:

@@

where I have defined $w(x|\psi) \coloneqq b^{\textrm{(ind)}}(x|\psi) - d^{\textrm{(ind)}}(x|\psi)$, the (Malthusian) `fitness' of the phenotype $x$. To track population numbers and trait frequencies, let us define as before, the total population size and population frequency as

@@

Where I have used the definition of $N_K(t)$ and assumed that integrals and derivatives commute in the second line. Substituting \eqref{PDE_for_kimura}, we now obtain

@@

where I have used the definition of $p(x,t)$ in the second line. Using \eqref{mean_fitness_for_kimura} and rearranging the terms gives us:

@@

This is a continuous version of the replicator-mutator equation when each $x$ is viewed as a strategy. It also yields Kimura's continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb

@@

where I have used the Fubini-Tonnelli theorem to go from the first step to the second, and have made the substitution $u = y-z$ to go from the third to the fourth step. We then note that since $m$ is a kernel, it satisfies $\int\limits_{\mathbb{R}} m(u)du = 1$, and \eqref{convolution_integral_intermediate} therefore becomes $\int\limits_{\mathbb{R}} Q(y|\psi)dy = N_K(t)$. Substituting this in \eqref{cts_replicator_mutator}, we have

@@

is the statistical covariance of the quantity $f$ with the Malthusian fitness function (Importantly, just like in the Price equation, this is an \emph{analogy} - Everything here is deterministic). The second term, which I will denote by

@@

Under these assumptions, if we supply an initial condition $\psi(x,0) = N_{K}(0)\delta_{y_0}$ for some constants $N_K(0) > 0$ and $y_0 \in \mathcal{T}$ (meaning we start with a completely monomorphic population of size $N_K(0)$ in which all individuals have trait value $y_0$), it is reasonable to assume that the population remains sufficiently well clustered for some (possibly small) time $t>0$ that we can continue to approximate the distribution $\psi(x,t)$ as a Dirac Delta mass $N_{K}(t)\delta_{y(t)}$ that is moving across the trait space in a deterministic manner dictated by a function $y(t)$ (to be found). Note that we have $p(x,t) = \delta_{y(t)}$, $\overline{x}(t) = y(t)$, and $\overline{w}(t) = 0$. Thus, from equation \eqref{cts_price}, we have

@@

where I have used $\overline{x}(t) = y(t)$. We can define the shorthand $B(y) =\int\limits_{\mathcal{T}}(x-y(t))^2p(x,t)dx = \int\limits_{\mathcal{T}}(x-\overline{x}(t))^2p(x,t)dx$ to obtain:

@@

\section{Stochastic trait frequency dynamics in the infinite-dimensional case}
In chapter \ref{chap_BD}, I also derived SDEs for the trait frequency dynamics of the complete stochastic case using It\^o's formula. Doing the same for quantitative traits in our framework is tricky because it requires us to find a version of It\^o's formula that holds for SPDEs of the form \eqref{functional_langevin}. The formulation of It\^o formulas and/or an infinite-dimensional stochastic calculus for general function (measure) valued stochastic processes is an active area of research in pure mathematics~\citep{da_prato_stochastic_2014}, and without further information on the nature of the domain $\mathcal{M}_{K}(\mathcal{T})$, it is not clear (to me at least) whether an It\^o's formula exists for our case in general. However, it bears noting that the `intuitive' It\^o's formula one would expect does indeed hold true for broad classes of Hilbert space valued processes~\citep{prevot_concise_2007,da_prato_stochastic_2014,liu_stochastic_2015}. If $\mathcal{A}^{\pm}(x | \phi)$ in \eqref{functional_FPE} are Gaussian functions, then it has been proven~\citep{week_white_2021} that the SDEs `work as expected' if we take $m \to \infty$ in \eqref{nD_eqn_for_frequencies} (see section \ref{sec_fun_theorems} in chapter \ref{chap_unification}). However, carrying out a general derivation is beyond the scope of this thesis.

@@

We can also formally carry out a functional analogue of the weak noise expansion as we did in chapter \ref{chap_BD}. Assume that $\psi(x,t)$ is the deterministic trajectory obtained as the solution to \eqref{deterministic_traj}. We introduce a new process $\{\zeta^{(s)}\}_{s \geq 0}$ which measures the fluctuations of $\phi^{(t)}$ from the deterministic trajectory $\psi(x,t)$. More precisely, let us introduce the new variables:
